---
title: Using `pool()` for multiple imputation in `mice` package
date: \today
author:
output:
  html_document:
    css: "C:/Users/belai/Documents/JBConsulting/GitHub/biostatistics/biostatistics.ca/theme.css"
    #toc: TRUE
    #toc_float : TRUE
    #toc_depth : 2
    header_includes:
      - \usepackage{amsmath, amssymb, amsthm, mathtools}
      - \newcommand{\training}{\mathcal{X} \times \mathcal{Y}}
      - \newcommand{\defeq}{:=}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, class.source = "rcode", class.output = "rcode")
```

By Justin Bélair &copy; | Biostatistician at [JB Statistical Consulting](www.justinbelair.ca)

[Find the RMarkdown Notebook on Github and Run the Code Yourself!](https://github.com/JB-Statistical-Consulting/biostatistics/tree/main/biostatistics.ca)

![](C:/Users/belai/Documents/JBConsulting/GitHub/biostatistics/biostatistics.ca/Justin-35(1).jpg){height=50%}

\newpage

# Multiple Imputation

Multiple Imputation (MI) is the cutting-edge statistical approach to dealing with missing data. Unfortunately, many often forego this technique because it requires substantial statistical and programming knowledge.

We'll show you that it need not be complicated for most of the analyses.


## Example 

We load our data and see there are missing values.

```{r, include = FALSE}
library(mice)
library(dplyr)
```


```{r}
# We look at our data

data.analysis <- read.csv("toy_data_binary_outcome_w_miss.csv")

str(data.analysis)

```

We are interested in estimating the effect of the Exposure on the **Binary_Outcome** variable, adjusted for **Age**, **Gender** and **Education**. We choose the logistic regression model.

```{r}
glm(Binary_Outcome ~ Exposure + Age + Gender + Education, data = data.analysis,
    family = binomial(link = "logit"))
```

Notice that `R` defaults to complete-case analysis, that is, any row with a missing value is removed. We see in the model output that 96 of the 247 observations were deleted due to missingness ($96/247 = 38.9\%$) - we cannot afford such loss of information. All the more considering that not all columns are missing from the deleted rows!

## Multiple imputation strategy :

1. We use the mice function to multiply impute, say `m=5` times. 
2. We call our logistic regression model on the multiply imputed object returned from our call to `mice` in a `with()` statement.
3. We use `pool()` function call. This will correctly "pool" the 5 imputations into a single model analysis.
4. We analyze the output, which has standard errors corrected for the uncertainty of the imputation^[See [van Buuren, Flexible Imputation of Missing Data](https://stefvanbuuren.name/fimd/) for more information].

```{r}
imp <- mice(data.analysis, m = 5, print = FALSE)

logistic.regression <- with(imp, 
                            glm(Binary_Outcome ~ Exposure + Age + Gender + Education, data = data.analysis, family = binomial(link = "logit"))
                            )

summary(pool(logistic.regression))

```

We get an output that allows us to do inference on model parameters, as usual and we find a significant effect of the **Exposure** on our **Binary_Outcome**!