---
title: Sample Size Estimates Done Via Power Calculation Are Uncertain
date: \today
author:
output:
  html_document:
    #toc: TRUE
    #toc_float : TRUE
    #toc_depth : 2
    header_includes:
      - \usepackage{amsmath, amssymb, amsthm, mathtools}
      - \newcommand{\training}{\mathcal{X} \times \mathcal{Y}}
      - \newcommand{\defeq}{:=}
---
[Find the RMarkdown Notebook on Github and Run the Code Yourself!](https://github.com/JB-Statistical-Consulting/biostatistics/tree/main/biostatistics.ca)

\newpage

## Introduction

I often see the results of power calculations used as if there is no uncertainty around the required sample size number obtained. This is a misconception, made almost invisible by the fact that traditional software tools like G-Power, SPSS, etc. don't provide any hint of the uncertainty around the sample size number they provide.

In short, estimates of the anticipated effect size from the study are by definition uncertain, or else the study wouldn't be needed - this uncertainty is transferred to the sample size number obtained via a power calculation.

Let's give a few quick examples.

## A Simple Example - the t-test

With an effect-size of 0.5, we need ~64 patients for a two-sample t-test with common standard deviation 1 to have 80% power at the 5% significance level.

```{r}
delta <- 0.5
sd <- 1

n <- power.t.test(delta = delta, sd = sd, sig.level = 0.05, power = 0.8)$n

n
```
### The t-test with an uncertain delta parameter

In practice, the delta is uncertain. Let's suppose there is uncertainty of about 5\% of the estimated value. For now, we assume the standard deviation is also known.

```{r}
delta.range <- c(1-0.05, 1, 1+0.05)*delta
n.delta.range <- sapply(delta.range, \(x) power.t.test(delta = x, sd = sd, sig.level = 0.05, power = 0.8)$n)

n.delta.range
```
We see that a range of ~58 to ~71 patients would be compatible with a small uncertainty around the estimate of delta.

### The t-test with an uncertain delta and standard deviation parameters

Let's add uncertainty around the standard deviation as well, using the same approach.

```{r}
sd.range <- c(1+0.05, 1, 1-0.05)*sd

n.delta.sd.range <- mapply(FUN = function(x,y){power.t.test(delta = x, sd = y, sig.level = 0.05, power = 0.8)$n},
                           delta.range, sd.range)

n.delta.sd.range
```
This additional uncertainty widens the range from ~53 to ~78 patients. The $n=64$ patients we found without considering the uncertainty inherent in estimating the effect-size and the standard deviation before hand could seriously mislead us if we do not account for it. 

## Conclusion

Be vigilant! Test a range of plausible values - you might be shocked by what you see. Your study might might heavily dependent on a false sense of confidence around the parameter values used to estimate the required sample size. 

Using a range of estimates for the effect-size and correspondingly determining a range of plausible sample sizes might give you a better understanding of the power you have to detect interesting effects.

Even better, consult a statistician to help you navigate the uncertainty in your study design. 

If you need help with statistics, don't hesitate to reach out to me at [JB Statistical Consulting](www.justinbelair.ca).
