---
title: 
output: pdf_document
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage[colorlinks=true, urlcolor=blue, allbordercolors={blue}, pdfborderstyle={/S/U/W 1}]{hyperref}
  - \usepackage{fancyhdr}
  - \fancypagestyle{plain}{\pagestyle{fancy}}
  - \pagestyle{fancy}
  - \fancyhead[LO,LE]{\href{https://justinbelair.ca/?utm_source=github&utm_medium=pdf&utm_campaign=kronecker_product}{JB Statistical Consulting}}
  - \fancyhead[RO,RE]{\href{https://www.biostatistics.ca/?utm_source=github&utm_medium=pdf&utm_campaign=kronecker_product}{biostatistics.ca}}
  - \renewcommand{\footrulewidth}{0.4pt}
  - \fancyfoot[LO,LE]{Visit the biostatistics.ca \href{https://www.biostatistics.ca/github-library/?utm_source=github&utm_medium=pdf&utm_campaign=kronecker_product}{\underline{\textbf{Github Library}}} for a full collection of our FREE R Markdown code notebooks.}
  - \fancyfoot[CO,CE]{}
---

# Kronecker Product

The Kronecker product is a special operation on matrices (arrays). It is defined as follows.

If $A$ is an $m \times n$ matrix and $\mathbf{B}$ is a $p \times q$ matrix, then the Kronecker product $A \otimes B$ $pm \times qn$ block matrix:

$$A \otimes \mathbf{B} = \begin{bmatrix} 
a_{11}\mathbf{B} & \cdots & a_{1n}\mathbf{B} \\ 
\vdots & \ddots & \vdots \\ 
a_{m1}\mathbf{B} & \cdots & a_{mn}\mathbf{B} 
\end{bmatrix},$$

where $a_{ij}$ is the $(i,j)$-th entry of $A$ and $\mathbf{B}$ is the matrix $B$. More explicitly, let's use:

- $A = \begin{bmatrix} 
1 & 2 & 3 \\ 
4 & 5 & 6
\end{bmatrix}$, a 2×3 matrix, and 
$B = \begin{bmatrix} 
1 & 2 \\ 
3 & 4
\end{bmatrix}$, a 2×2 matrix. Then, the Kronecker product $A \otimes B$ is computed as follows:

$$A \otimes \mathbf{B} = 
  \begin{bmatrix}
    1 & 2 & 3 \\
    4 & 5 & 6
  \end{bmatrix} \otimes 
  \begin{bmatrix}
    1 & 2 \\
    3 & 4
  \end{bmatrix} = \begin{bmatrix} 1
        \begin{bmatrix}
          1 & 2 \\ 3 & 4 
        \end{bmatrix} & 2
        \begin{bmatrix} 
          1 & 2 \\ 3 & 4 
        \end{bmatrix} & 3
        \begin{bmatrix} 
          1 & 2 \\ 3 & 4 
        \end{bmatrix} \\ 
        4
        \begin{bmatrix} 
          1 & 2 \\ 3 & 4 
        \end{bmatrix} & 5
        \begin{bmatrix} 
          1 & 2 \\ 3 & 4 
        \end{bmatrix} & 6
        \begin{bmatrix} 
          1 & 2 \\ 3 & 4 
        \end{bmatrix}
    \end{bmatrix} \\
    = \begin{bmatrix}
          1 & 2 & 2 & 4 & 3 & 6 \\
          3 & 4 & 6 & 8 & 9 & 12 \\
          4 & 8 & 5 & 10 & 6 & 12 \\
          12 & 16 & 15 & 20 & 18 & 24
      \end{bmatrix}$$

In R, this is easy to implement

```{r}
A <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 2, byrow = TRUE)
B <- matrix(c(1, 2, 3, 4), nrow = 2, byrow = TRUE)
A %x% B
```
## Kronecker Product Use-Case From My Consulting Practice

Suppose I want to simulate a quasi-poisson variable based on given mean parameters $\lambda$ and $\phi$. For a quasi-poisson, the variance is equal to $\phi \cdot \lambda$. Suppose I have a matrix of means $\lambda$, that are determined by a combination of 2-variables. I also have a set of parameters $\phi$. Now, I want all the possible combinations of $\lambda \cdot phi$, i.e all the possible variances. I can create this using the Kronecker product. The matrix is given by:

```{r}
lambda <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 2, byrow = TRUE)
phi <- c(3, 4)
phi %x% lambda
```
Now, I can iterate over these variance parameters for my analysis.